quollr: An R Package for Visualizing 2D Models from Nonlinear Dimension Reductions in High Dimensional Space

### 2024-06-04 paper-quollr.Rmd
Plan for paper

---
title: "quollr: An R Package for Visualizing 2D Models from Nonlinear Dimension Reductions in High Dimensional Space"
abstract: >
  Non-Linear Dimension Reduction (NLDR) techniques have emerged as powerful tools to visualize high-dimensional ($p-D$) data in low-diemnsioanl space. However, their complexity and hyperparameter choices may lead to distrustful or misleading results. The R package \CRANpkg{quollr} is developed as a new tool to help to determine which method, which hyperparameter choice provide the most accurate representation of $p-D$ data. Clustering data from \CRANpkg{cardinalR} package, is used to illustrate the algorithm and its use within the package.
draft: true
author:  
  - name: Jayani P.G. Lakshika
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email:  \email{jayani.piyadigamage@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
creative_commons: CC BY
date: "`r Sys.Date()`"
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
output: 
 rjtools::rjournal_web_article:
    css: "style.css"
    keep_md: true
bibliography: RJreferences.bib
editor_options: 
  chunk_output_type: inline
---

@@

This paper presents the R package, `quollr` which introduce a new tool in determining which NLDR technique and which hyperparameter choice gives most accurate representation of $p-D$ data. The methodology of the algorithm is explained in *cite the methodology paper*. Furthermore, the `quollr` package enables users to perform hexagonal binning [@dan2023], resulting in the generation of regular hexagons. The software is available from the Comphrehensive R Archive Network (CARN) at [https://CRAN.R-project.org/package=quollr](https://CRAN.R-project.org/package=quollr).

@@

The following demonstration of the package's functionality assumes `quollr` has been loaded. We also want to load the built-in data sets `s_curve_noise_training` and `s_curve_noise_umap`.

The mains steps for the algorithm can be executed by the main function `fit_highd_model()`, or can be run separately for more flexibility. When constructing the 2D model, the user can choose either to fit the 2D model with hexagonal bin centroids or bin means using `is_bin_centroid` argument.

```{r, echo=TRUE, eval=FALSE}
model_object <- fit_highd_model( training_data = s_curve_noise_training, 
                                 nldr_df_with_id = s_curve_noise_umap_scaled, 
                                 x = "UMAP1", y = "UMAP2", 
                                 num_bins_x = num_bins_x_s_curve, 
                                 num_bins_y = num_bins_y_s_curve, 
                                 x_start = NA, y_start = NA, 
                                 buffer_x = NA, buffer_y = NA, 
                                 hex_size = hex_size_s_curve,
                                 is_bin_centroid = TRUE,
                                 is_rm_lwd_hex = FALSE, 
                                 benchmark_to_rm_lwd_hex = NA, 
                                 col_start_2d = "UMAP", 
                                 col_start_highd = "x")
```

@@

```{r}

## Decide by looking at MSE plot
num_bins_x_s_curve <- 8
num_bins_y_s_curve <- 14
hex_size_s_curve <- 0.1
hb_obj_s_curve <- hex_binning(data = s_curve_noise_umap_scaled, x = "UMAP1", 
                      y = "UMAP2", num_bins_x = num_bins_x_s_curve, 
                      num_bins_y = num_bins_y_s_curve, x_start = NA, y_start = NA, 
                      buffer_x = NA, buffer_y = NA, hex_size = hex_size_s_curve, col_start = "UMAP")
## Data set with all possible centroids in the hexagonal grid
all_centroids_df <- as.data.frame(do.call(cbind, hb_obj_s_curve$centroids))
## Generate all coordinates of hexagons
hex_grid <- as.data.frame(do.call(cbind, hb_obj_s_curve$hex_poly))
## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_s_curve$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
## remove low-density hexagons
hex_grid_with_counts <- dplyr::left_join(hex_grid, counts_df, by = c("hex_poly_id" = "hb_id"))

centroid_plot <- ggplot(data = all_centroids_df, aes(x = c_x, y = c_y)) +
  geom_point(color = "#33a02c") +
  coord_fixed() +
  theme_void() +
  theme(plot.title = element_text(size = 7, hjust = 0.5,
vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6))

full_grid <- ggplot(data = hex_grid, aes(x = x, y = y)) +
  geom_polygon(fill = "#ffffff", color = "black", aes(group = hex_poly_id)) +
  geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "#33a02c") +
  coord_fixed() +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5,
vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6))

std_pts_grid <-  ggplot(data = hex_grid_with_counts, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_point(data = s_curve_noise_umap_scaled, aes(x = UMAP1, y = UMAP2), color = "black", alpha = 1) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  coord_fixed() +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5,
vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6))

umap_data_with_hb_id <- as.data.frame(do.call(cbind, hb_obj_s_curve$data_hb_id))

model_object <- fit_highd_model( training_data = s_curve_noise_training, 
                                 nldr_df_with_id = s_curve_noise_umap_scaled, 
                                 x = "UMAP1", y = "UMAP2", 
                                 num_bins_x = num_bins_x_s_curve, 
                                 num_bins_y = num_bins_y_s_curve, 
                                 x_start = NA, y_start = NA, 
                                 buffer_x = NA, buffer_y = NA, 
                                 hex_size = hex_size_s_curve,
                                 is_rm_lwd_hex = FALSE, 
                                 benchmark_to_rm_lwd_hex = NA, 
                                 col_start_2d = "UMAP", 
                                 col_start_highd = "x")
df_bin_centroids_s_curve <- model_object$df_bin_centroids
df_bin_s_curve <- model_object$df_bin

## Triangulate bin centroids
tr1_object_s_curve <- tri_bin_centroids(df_bin_centroids_s_curve, x = "c_x", y = "c_y")
tr_from_to_df_s_curve <- gen_edges(tri_object = tr1_object_s_curve)
## Compute 2D distances
distance_s_curve <- cal_2d_dist(tr_coord_df = tr_from_to_df_s_curve, 
                             start_x = "x_from", start_y = "y_from", 
                             end_x = "x_to", end_y = "y_to", 
                             select_vars = c("from", "to", "distance"))

## To find the benchmark value
benchmark_s_curve <- find_lg_benchmark(distance_edges = distance_s_curve, distance_col = "distance")
benchmark_s_curve <- 0.5

trimesh_s_curve <- ggplot(df_bin_centroids_s_curve, aes(x = c_x, y = c_y)) +
  geom_trimesh() +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3) +
  labs(colour = "")

trimesh_colored_s_curve_umap <- vis_lg_mesh(distance_edges = distance_s_curve, benchmark_value = benchmark_s_curve,
tr_coord_df = tr_from_to_df_s_curve, distance_col = "distance")

trimesh_colored_s_curve_umap <- trimesh_colored_s_curve_umap +
  theme_linedraw() +
  theme(legend.position = "bottom", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3) +
  labs(colour = "")

trimesh_removed_s_curve_umap <- vis_rmlg_mesh(distance_edges = distance_s_curve, benchmark_value = benchmark_s_curve, tr_coord_df = tr_from_to_df_s_curve, distance_col = "distance")
trimesh_removed_s_curve_umap <- trimesh_removed_s_curve_umap +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'c', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

trimesh_removed_s_curve_umap_0.3 <- vis_rmlg_mesh(distance_edges = distance_s_curve, benchmark_value = 0.3, tr_coord_df = tr_from_to_df_s_curve, distance_col = "distance")
trimesh_removed_s_curve_umap_0.3 <- trimesh_removed_s_curve_umap_0.3 +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

trimesh_removed_s_curve_umap_0.8 <- vis_rmlg_mesh(distance_edges = distance_s_curve, benchmark_value = 0.8, tr_coord_df = tr_from_to_df_s_curve, distance_col = "distance")
trimesh_removed_s_curve_umap_0.8 <- trimesh_removed_s_curve_umap_0.8 +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

df_all_s_curve <- dplyr::bind_cols(s_curve_noise_training |> dplyr::select(-ID), umap_data_with_hb_id)

# show_langevitour(df_all_s_curve, df_bin_s_curve, df_bin_centroids_s_curve, benchmark_s_curve, distance_s_curve, "distance", col_start = "x")

### Define type column
df <- df_all_s_curve |>
  dplyr::select(tidyselect::starts_with("x")) |>
  dplyr::mutate(type = "data") ## original dataset

df_b <- df_bin_s_curve |>
  dplyr::filter(hb_id %in% df_bin_centroids_s_curve$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean
## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b <- df_b[match(df_bin_centroids_s_curve$hexID, df_b$hb_id),] |>
  dplyr::select(-hb_id)
df_exe <- dplyr::bind_rows(df_b, df)
## Set the maximum difference as the criteria
distance_df_small_edges <- distance_s_curve |>
  dplyr::filter(distance < benchmark_s_curve)
## Since erase brushing is considerd.
tour <- langevitour::langevitour(df_exe[1:(length(df_exe)-1)],
                         lineFrom = distance_df_small_edges$from,
                         lineTo = distance_df_small_edges$to,
                         group = df_exe$type, pointSize = append(rep(0, NROW(df_b)), rep(1, NROW(df))),
                         levelColors = c("#6a3d9a", "#33a02c"))
```

## Constructing the 2D model

Constructing the 2D model mainly contains (1) preprocessing, (2) binning NLD data, (3) obtain bin centroids/ means, and (4) triangulate bin centroids/ means.

### Preprocessing
<!--Discuss scaling the NLDR data-->

The function `gen_scaled_data()` is used to prepare the NLDR data to fit within the bounds required for regular hexagonal binning, ensuring effective visualization. The user can set up the ratio of the width and height of the hexagon using the `hex_ratio` argument.

```{r, echo=TRUE, eval=FALSE}
s_curve_noise_umap_scaled <- gen_scaled_data(data = s_curve_noise_umap, 
                                             x = "UMAP1", y = "UMAP2", 
                                             hex_ratio = NA)

<!-- add plot before and after scaling NLDR data-->
<!-- also can discuss about the tuning hex_ratio-->

### Binning data

In binning data step, the objective is to perform hexagonal binning and assign NLDR data to regualr hexagons.

**Step 1: Compute hexagonal grid configurations**

```{r, echo=TRUE, eval=FALSE}
bin_list <- calc_bins(data = s_curve_noise_umap_scaled, x = "UMAP1", 
                      y = "UMAP2", hex_size = 0.1, buffer_x = NA, buffer_y = NA)

num_bins_x <- bin_list$num_x
num_bins_y <- bin_list$num_y
```

The mains steps for the hexagonal binning algorithm can be executed by the main function `hex_binning()`, or can be run separately for more flexibility.
<!-- add about main function for hexagonal binning-->

```{r, echo=TRUE, eval=FALSE}
hb_obj <- hex_binning(data = s_curve_noise_umap_scaled, x = "UMAP1", 
                      y = "UMAP2", num_bins_x = num_bins_x, 
                      num_bins_y = num_bins_y, x_start = NA, y_start = NA, 
                      buffer_x = NA, buffer_y = NA, hex_size = NA, 
                      col_start = "UMAP")

```{r, out.width="100%", fig.cap = "Illustration of key steps of the hexagonal bining algorithm: (a) first, all hexagonal bin centroids are created; (b) second, full hexagon grid is created; and (c) finally, assign the data to hexagons. The centroids are used to generate hexagonal coordinates. Once the full hexagonal grid is generated, by computing the 2D euclidean distance from each NLDR points to hexagonal bin centroids, the nearest NLDR points are allocated to the nearest hexagoanl bin."}

centroid_plot + full_grid + std_pts_grid +
  plot_annotation(tag_levels = 'a') +
  plot_layout(ncol = 3) &
  theme(plot.tag = element_text(size = 8))
```

**Step 2: Generate the hexagonal bin centroids**

```{r, echo=TRUE, eval=FALSE}
centroid_list <- gen_centroids(data = s_curve_noise_umap_scaled, x = "UMAP1", 
                               y = "UMAP2", num_bins_x = num_bins_x,
                               num_bins_y = num_bins_y, x_start = NA,
                               y_start = NA, buffer_x = NA, buffer_y = NA, 
                               hex_size = NA)

all_centroids_df <- as.data.frame(do.call(cbind, centroid_list))
```

**Step 3: Create the hexagonal grid**

```{r, echo=TRUE, eval=FALSE}
all_hex_coordinates_list <- gen_hex_coord(centroids_df = all_centroids_df,
                                            hex_size = NA)
```

**Step 4: Assign NLDR data to hexagons**

```{r, echo=TRUE, eval=FALSE}
df_with_hb_id_list <- assign_data(data = s_curve_noise_umap_scaled |> 
                                    dplyr::select(-ID), 
                                  centroid_df = all_centroids_df, 
                                  col_start = "UMAP")
umap_data_with_hb_id <- as.data.frame(do.call(cbind, df_with_hb_id_list))
```

```{r, echo=TRUE, eval=FALSE}
std_counts_list <- compute_std_counts(data_hex_id = df_with_hex_id)
counts_df <- as.data.frame(do.call(cbind, std_counts_list))
```

### Obtain bin centroids/ means

To obtain hexagonal bin centroids, `extract_hexbin_centroids()`is used. The function `extract_hexbin_mean()` is used to obtain average within each hexagon.

```{r, echo=TRUE, eval=FALSE}
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
```

```{r, echo=TRUE, eval=FALSE}
df_bin_means <- extract_hexbin_mean(nldr_df_with_hex_id = umap_with_hb_id,
                                        counts_df = counts_df)
```

### Triangulate bin centroids/ means

The `tri_bin_centroids()` is used to do triangulate bin centroids/ bin means. Then, `gen_edges()` compute the edges to obtain the triangular mesh which is the 2D model.

```{r, echo=TRUE, eval=FALSE}
tr1_object <- tri_bin_centroids(hex_df = df_bin_centroids, x = "c_x", y = "c_y")
tr_from_to_df <- gen_edges(tri_object = tr1_object)
```

@@

The $p$-D model is generated by `avg_highd_data()` by passing the $p$-D data.

```{r, echo=TRUE, eval=FALSE}
df_all <- dplyr::bind_cols(s_curve_noise_training |> dplyr::select(-ID),
                           umap_data_with_hb_id)

df_bin <- avg_highd_data(data = df_all, col_start = "x")
```

@@

Mainly there are two model parameter need to consider: benchmark value to remove the low-density hexagons, and benchmark value to remove the long edges. `find_low_dens_hex()` is used to find the hexagons which contains less number of points by considering the density of their neighboring points as well. In here, user can first decide which are the low-density hexagons and pass them to this function to check whether these removal of low-density hexagons can affect the model fit by looking at the the neighbors. On the other hand, `find_lg_benchmark()` function is used to compute the threshold for removing long edges.

```{r, echo=TRUE, eval=FALSE}
df_bin_centroids_low <- df_bin_centroids |> dplyr::filter(std_counts <= 0.43)

low_dens_hex <- find_low_dens_hex(df_bin_centroids_all = df_bin_centroids, 
                  num_bins_x = num_bins_x, 
                  df_bin_centroids_low = df_bin_centroids_low)
```

```{r, echo=TRUE, eval=FALSE}
distance_df <- cal_2d_dist(tr_coord_df = tr_from_to_df, start_x = "x_from", 
                           start_y = "y_from", end_x = "x_to", end_y = "y_to",
                           select_vars = c("from", "to", "distance"))

benchmark_val <- find_lg_benchmark(distance_edges = distance_df, distance_col = "distance")
```

@@

`predict_emb()` function is used to predict 2D embedding for a new $p$-D data point using the fitted model. This function is useful to predict 2D embedding irrespective of the NLDR technique.

```{r}
## Prediction for training data
pred_emb_list_training <- predict_emb(test_data = s_curve_noise_training, 
                             df_bin_centroids = df_bin_centroids_s_curve, 
                             df_bin = df_bin_s_curve, type_NLDR = "UMAP")
pred_df_training <- as.data.frame(do.call(cbind, pred_emb_list_training))

pred_points_training <- ggplot(data = s_curve_noise_umap_scaled, aes(x = UMAP1, y = UMAP2)) +
  geom_point() +
  geom_point(data = pred_df_training, aes(x = pred_UMAP_1, y = pred_UMAP_2), colour = "#de2d26") +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

## Prediction for test data
pred_emb_list_test <- predict_emb(test_data = s_curve_noise_test, 
                             df_bin_centroids = df_bin_centroids_s_curve, 
                             df_bin = df_bin_s_curve, type_NLDR = "UMAP")
pred_df_test <- as.data.frame(do.call(cbind, pred_emb_list_test))

pred_points_test <- ggplot(data = s_curve_noise_umap_scaled, aes(x = UMAP1, y = UMAP2)) +
  geom_point() +
  geom_point(data = pred_df_test, aes(x = pred_UMAP_1, y = pred_UMAP_2), colour = "#de2d26") +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```{r, out.width="100%", fig.cap = "Predicted 2D embedding overlaid with UMAP embedding of S-curve training data: (a) prediction of 2D embeddings for S-curve training data, and (b) prediction of 2D embeddings for S-curve test data. Once the high-D model is generated, first, for the new data points the nearest high-D mappings of hexagonal bin centroids find by computing Euclidean distance. Then, mapping back the 2D hexagonal bin centroids gives the prediction for the new data points."}

pred_points_training + pred_points_test +
  plot_layout(ncol = 2) &
  theme(plot.tag = element_text(size = 8))
```

@@

As a Goodness of fit statistics for the model, `gen_summary()` is used to compute residuals and MSE. For the fitted model, how residuals and MSE is changing is shown in Figure 2.

```{r, echo=TRUE}
gen_summary(test_data = s_curve_noise_training, prediction_df = pred_df_training,
df_bin = df_bin_s_curve, col_start = "x")
```

```{r, echo=TRUE}
gen_summary(test_data = s_curve_noise_test, prediction_df = pred_df_test,
df_bin = df_bin_s_curve, col_start = "x")
```

<!--add a plot with total bins Vs MSE, non-empty bins Vs Error-->

@@

The package provides four basic visualizations which includes three visualizations regarding 2D model (static vis) and one is regrading $p$-D model (dynamic vis). Each visualization can be done their own functions which describes in this section.

<!--add trimesh, colored trimesh, removed edge trimesh, screenshots from langevitour with the model and data-->

### 2D model visualization

<!--2D model vis: triangular mesh, colored trimesh, removed edge trimesh-->
To visualize the 2D model, mainly three functions are used. As shown in Figure \@ref(fig:mesh-plots)a, `geom_trimesh()` to visualize the triangular mesh by adding a new layer to `ggplot()`. After identifying benchmark value to remove long edge, `vis_lg_mesh()` is used to visualize the triangular mesh by coloring the small and long edges. As shown in Figure \@ref(fig:mesh-plots)b, the small and long edges are colored by black and red respectively. Following this, `vis_rmlg_mesh()` is used to visualize smoothed 2D model which is the 2D model after removing the long edges (see Figure \@ref(fig:mesh-plots)c). In `vis_lg_mesh()` and `vis_rmlg_mesh()`, `benchmark_value` argument controls the edge removal in 2D. Using small value of `benchmark_value`, will produce a triangular mesh with missing data structure; for example `benchmark_value = 0.3` shows two clusters rather than continuous structure, while `benchmark_value = 0.8` creates long edges and mislead the data structure in $p-D$ space.

```{r mesh-plots, out.width="100%", fig.cap="Visualization of model generated for UMAP embedding for S-curve data in 2D space: (a) triangular mesh, (b), triangular mesh colored by edge type, and (c) triangular mesh after removing the long edges. The edges which has the length greater than $0.5$ are assigned as long edges. To obtain smooth representation in 2D space, the long edges are removed."}

trimesh_s_curve + trimesh_colored_s_curve_umap + trimesh_removed_s_curve_umap +
  plot_layout(ncol = 3) &
  theme(plot.tag = element_text(size = 8))
```

```{r mesh-diffplots, out.width="100%", fig.cap="Visualization of model generated for UMAP embedding for S-curve data in 2D space with different benchmark value applied to remove long edges: (a) benchmark_value = 0.3, and (b) benchmark_value = 0.8. Small benchmark values will generate misleading data structures in 2D space, while; large bechmark value will generate more long edges in $p-D$ and miinterpret the $p-D$ data structure."}

trimesh_removed_s_curve_umap_0.3 + trimesh_removed_s_curve_umap_0.8 +
  plot_layout(ncol = 2) &
  theme(plot.tag = element_text(size = 8))
```

### p-D model visualization

<!--highD model vis: langevitour with the model-->

The $p$-D model overlaid on data is visualized by the function `show_langevitour()`. This is a useful dynamic visualization result to validate visually whether the model fits the data well or not.

```{r, echo=TRUE, eval=FALSE}
tour <- show_langevitour(df = df_all, df_b = df_bin, 
                 df_b_with_center_data = df_bin_centroids, benchmark_value = 0.5, 
                 distance = distance_df, distance_col = "distance", 
                 use_default_benchmark_val = FALSE, col_start = "x")
```

<!--need to add screenshots of the frames and add the link of the video-->
```{r tour-s-curve}
tour
```

@@

<!--in data processing, minus values, NA, Inf values for aspect_ratio or hex_ratio-->
<!-- to check length of list objects-->
<!-- add tests to find NAs, inf values are in NLDR data-->

@@

We illustrate the use of the algorithm by applying it to `clusters_different_shapes_diff_num_points` dataset in `cardinalR` package. The tSNE is used in the example for the demonstration (see Figure \@ref(fig:tsne-clust)).

<!--select a dataset from cardinalR-->
<!-- diff shape clusters-->
<!-- talk with best fit-->

```{r, echo=FALSE}
clusters_different_shapes_diff_num_points <- function(sample_size = 400, with_seed = NULL, cluster_size_vec = c(50, 50, 50, 50, 100, 100), num_gussian_clusters = 4, num_non_gaussian_clusters = 2, cluster_sd_gau = 0.05, cluster_sd_non_gau = 0.1, num_dims = 7, a = 2, b = 4) {

  # To check the seed is not assigned
  if (!is.null(with_seed)) {
    set.seed(with_seed)
  }

  num_clusters <- num_gussian_clusters + num_non_gaussian_clusters

  ## Generate Gaussian clusters

  # Create a vector of possible values (0 and 1)
  values <- c(0, 1)

  # Create an expanded grid with 0's and 1's
  mean_val_grid <- tidyr::expand_grid(!!!setNames(rep(list(values), num_dims),
                                                  paste0("mean_dim", 1:num_dims)))

  # To select combinations for assigned number of clusters

  mean_val_grid_gau <- mean_val_grid |>
    dplyr::slice_sample(n = num_gussian_clusters)

  mean_val_grid_non_gau <- mean_val_grid |>
    dplyr::slice_sample(n = num_non_gaussian_clusters)

  # To generate empty tibble
  column_names <- paste0(rep("x", num_dims), 1:num_dims)
  df <- tibble(!!!setNames(rep(list(NULL), length(column_names)), column_names))

  for (i in 1:num_gussian_clusters) {

    # To filter the mean values for specific cluster
    mean_val_for_cluster <- mean_val_grid_gau |>
      dplyr::filter(dplyr::row_number() == i) |>
      unlist(use.names = FALSE)

    # Initialize an empty list to store the vectors with column
    # values
    dim_val_list <- list()

    for (j in 1:num_dims) {

      dim_val_list[[column_names[j]]] <- rnorm(cluster_size_vec[i], mean = mean_val_for_cluster[j],
                                               sd = cluster_sd_gau)

    }
    # To generate a tibble for a cluster
    df_gau_cluster <- tibble::as_tibble(dim_val_list)

    df <- dplyr::bind_rows(df, df_gau_cluster)

  }

  for (i in 1:num_non_gaussian_clusters) {

    phi <- runif(cluster_size_vec[(num_clusters - i)], max = 2*pi)
    rho <- sqrt(runif(cluster_size_vec[(num_clusters - i)]))

    # To filter the mean values for specific cluster
    presence_of_elipse_cluster <- mean_val_grid_non_gau |>
      dplyr::filter(dplyr::row_number() == i) |>
      unlist(use.names = FALSE)

    # Initialize an empty list to store the vectors with column
    # values
    dim_val_list_n <- list()

    for (j in 1:num_dims) {
      if(presence_of_elipse_cluster[j] == 1){
        dim_val_list_n[[column_names[j]]] <- sqrt(a)*rho*cos(phi) + b
        ## Surface of poolar coordinate
      } else {
        dim_val_list_n[[column_names[j]]] <- rnorm(cluster_size_vec[(num_clusters - i)], mean = 0,
                                                   sd = cluster_sd_non_gau)

      }

    }
    # To generate a tibble for a cluster
    df_non_gau_cluster <- tibble::as_tibble(dim_val_list_n)

    df <- dplyr::bind_rows(df, df_non_gau_cluster)

  }

  df

}
```

<!--need to add screen shots of the original data-->

```{r}
clust_df <- clusters_different_shapes_diff_num_points(sample_size = 1500, with_seed = NULL, cluster_size_vec = c(250, 150, 150, 150, 350, 450), num_gussian_clusters = 4, num_non_gaussian_clusters = 2, cluster_sd_gau = 0.2, cluster_sd_non_gau = 0.3, num_dims = 5, a = 2, b = 4)

langevitour::langevitour(clust_df)

clust_df <- clust_df |>
  dplyr::mutate(ID = row_number())
```

```{r}
tSNE_fit <- clust_df |>
  dplyr::select(-ID) |>
  dplyr::select(where(is.numeric)) |>
  Rtsne::Rtsne(perplexity = 39, pca = FALSE, pca_center = FALSE, normalize = FALSE, dims = 2)

tSNE_df <- tSNE_fit$Y |> 
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::mutate(ID = clust_df$ID)

names(tSNE_df) <- c("tSNE1", "tSNE2", "ID")

tSNE_df_plot <- tSNE_df |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.5, colour="#e41a1c", size = 0.5) +
  coord_equal() +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        axis.text = element_text(size = 5),
        axis.title = element_text(size = 7))
```

```{r tsne-clust, out.width="100%", fig.cap="tSNE representation of data with perplexity: 39. Is this representation accurately capture the structure in high dimensions?"}
tSNE_df_plot
```

## Selecting parameter values for hexagonal binning

We previously introduced three key parameters, *hex_size*, *x_start*, and *y_start*, in our explanation of the algorithm. While their best choice is unknown, a visualization tool enables their tuning such that user can select reasonable values.

Consider the relationships between *total number of bins*, *starting point*, and MSE. Increase of *total number of bins* usually reduces the MSE.

<!--hex_size, starting point (x_start, y_start)-->
<!--talk about to use bin centroids or bin means-->

>>>>

---
title: "quollr: An R Package for Visualizing 2D Models from Nonlinear Dimension Reductions in High Dimensional Space"
abstract: >
  Non-Linear Dimension Reduction (NLDR) techniques have emerged as powerful tools to visualize high-dimensional data in low-diemnsioanl space. However, their complexity and (hyper)parameter choices may lead to distrustful or misleading results. The R package \CRANpkg{quollr} is developed as a new tool to help to determine which method, which (hyper)parameter choice provide the most accurate representation of $p-D$ data. Clustering data from \CRANpkg{cardinalR} package, is used to illustrate the algorithm and its use within the package.
draft: true
author:  
  - name: Jayani P.G. Lakshika
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email:  \email{jayani.piyadigamage@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
creative_commons: CC BY
date: "`r Sys.Date()`"
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
output: 
 rjtools::rjournal_web_article:
    css: "style.css"
    keep_md: true
bibliography: RJreferences.bib
editor_options: 
  chunk_output_type: inline
---

@@

This paper presents the R package, `quollr` which introduce a new visual tool in determining which NLDR technique and which (hyper)parameter choice gives most accurate representation of high-dimensional data. The methodology of the algorithm is explained in *cite the methodology paper*. Furthermore, the `quollr` package enables users to perform hexagonal binning [@dan2023], resulting in the generation of regular hexagons. The software is available from the Comprehensive R Archive Network (CARN) at [https://CRAN.R-project.org/package=quollr](https://CRAN.R-project.org/package=quollr).

@@

@@

## Constructing the 2-D model

### Scaling the data

### Computing hexagon grid configurations

### Binning the data

### Indicating neighbors by line segments connecting centroids

@@

@@

@@

@@

@@

### $2\text{-}D$ model visualization

### $p\text{-}D$ model visualization

@@

<!--discuss a test with how a point in a intersection of two hexagonal going to evaluate-->

@@

<!--with prism data-->

# Discussion
